Measuring Directionality in Double Beta Decay and Neutrino Interactions with Kiloton-Scale Scintillation Detectors

### 2013-07-18 Direction.tex
Add new text from Matt and figure caption. --Lindley

\begin{figure*}[tbh]
        \begin{center}
        \subfigure{\includegraphics[scale=0.295]{graphs/hDirX_v1.pdf}}
        \subfigure{\includegraphics[scale=0.295]{graphs/hDirY_v1.pdf}}
        \subfigure{\includegraphics[scale=0.295]{graphs/hDirZ_v1.pdf}}
        \subfigure{\includegraphics[scale=0.295]{graphs/hVtxX_v1.pdf}}
        \subfigure{\includegraphics[scale=0.295]{graphs/hVtxY_v1.pdf}}
        \subfigure{\includegraphics[scale=0.295]{graphs/hVtxZ_v1.pdf}}
        \caption[]{We need a caption}
        \end{center}
\end{figure*}

WCSimAnalysis is a Water Cherenkov reconstruction package developed by Andrew Blake for the Long Baseline Neutrino Experiment (LBNE collaboration). It provides a framework for generic event cleaning, track reconstruction, and particle id, and it comes equipped with variety of pre-built algorithms. A collaboration between Iowa State, the University of Chicago, and Argonne National Laboratory has continued adding to the code, developing new track-fitting techniques for Water Cherenkov detectors built around the unique capabilities of microchannel plate photodetectors (MCPs): timing resolutions below 100 picoseconds and imaging resolutions below a centimeter. In this paper, we use only the simple, low-level reconstruction tools native to the package, though in future works we hope to implement more sophisticated techniques.

The simplest approach to vertex reconstruction, used in this paper, is commonly referred to as a “point fit”. It assumes that, to first order, the light was emitted a single point in space-time (x0,y0,z0,t0). In reality, the light is emitted along the extended, multi-scattered track of the electron. However, at the energies discussed in this paper, the extent of the electron track is small (a few cm) compared to the scale of the detector, which has a radius of 6 meters. It is also small compared with the vertex resolutions typical of experiments with this scale. As starting point, the assumption works well.

The first algorithm we use to find a seed vertex relies on exact numerical calculations of possible vertex candidates. Given a single point source, we need four constraints to solve for the four unknowns of the vertex. A collection of N random quadruplets of hits is selected and used to solve for N corresponding vertex candidates. Some of these solutions will produce nonsensical results due to optical scattering of the emitted light, slow decay constant of the scintillation, finite detector resolution, and the limitations of our assumption of a point source. Nonetheless, over the N unique quadruplets, we expect some of the solutions to lie near the true event vertex. In this paper we chose N to be 80, determined by trial and error. However, given many hundreds of hits, the number of possible 4-hit combinations is much larger and one can choose a different N to better suit one’s particular experimental context.

Once the 80 vertex candidates have been solved for, we test the goodness of each candidate and select the one that best fits the full ensemble of hits. Distances are calculated between each hit and the candidate point-source. From these distances, we calculate the transit time, given the speed of light in our scintillator medium. This transit time plus the t0 of the 4-vertex provides a predicted arrival time for each hit. The “time residual” is the calculated as difference between the predicted and the measured arrival time, given the vertex candidate. Since we assume a point-source in this paper, we use what we call the “simple time residual”, as distinct from the “extended time residual”, which assumes an extended, track-like geometry. The distribution of time residuals for each hit with respect to a vertex candidate has a width that is minimized when the hypothesis vertex is close to the true vertex. The t0 of the vertex candidate is left to float, and is adjusted to maximize the likelihood comparison with a Gaussian test function. The width of that gaussian test function is hand optimized to correspond with the width of the “best fit”. The vertex candidate with the largest maximum is selected to serve as the best candidate seed vertex. Once a single best seed vertex has been chosen, the same time-residual figure-of-merit is used in a MINUIT-based fit over a continuum of possible vertices in the neighborhood of the seed.

Once the vertex fitting is completed, we determine the direction by taking the centroid of all vectors pointing from the fitted event vertex and the hits on the detector. Since our initial time cut greatly enhances the purity of the Cherenkov light in this sample of hits and, since the Cherenkov light is directional (compared to the isotropic scintillation light), this fit provides a good measure of the track direction. This approach does not rely on the highly constrained Cherenkov geometry of the event, which contains further information. Future work will focus on better algorithms. Nonetheless, for a proof-of-principles, this method works sufficiently well..

>>>>

\begin{figure*}[tbh]
        \begin{center}
        \subfigure{\includegraphics[scale=0.295]{graphs/hDirX_v1.pdf}}
        \subfigure{\includegraphics[scale=0.295]{graphs/hDirY_v1.pdf}}
        \subfigure{\includegraphics[scale=0.295]{graphs/hDirZ_v1.pdf}}
        \subfigure{\includegraphics[scale=0.295]{graphs/hVtxX_v1.pdf}}
        \subfigure{\includegraphics[scale=0.295]{graphs/hVtxY_v1.pdf}}
        \subfigure{\includegraphics[scale=0.295]{graphs/hVtxZ_v1.pdf}}
        \caption[]{\label{fig:reco} (Top) Reconstructed direction, $(\frac{p_x}{|\vec{p}|}, \frac{p_y}{|\vec{p}|}, \frac{p_z}{|\vec{p}|})$, for the simulation of 1000 electrons (5MeV). In the simulation the electrons are produced along the X-axis, $\frac{\vec{p}}{|\vec{p}|}$ = (1,0,0), and originate from the center of the detector, $\vec{r}$ = (0,0,0). Detector diameter, $R$ = 6.5~m. Only photons with arrival time of $T<$ 34~ns are used in the reconstruction. The quantum efficiency of the bialkali photocathode is taken into account. (Bottom) Reconstructed vertex position, $(x,y,z)$, for the simulation of 1000 electrons (5~MeV). In the simulation the electrons are produced along the x-axis, $\frac{\vec{p}}{|\vec{p}|}$ = (1,0,0), and originate from the center of the detector, $\vec{r}$ = (0,0,0). Detector diameter, $R$ = 6.5~m. Only photons with arrival time of $T<$ 34~ns are used in the reconstruction. The quantum efficiency of the bialkali photocathode is taken into account. }
        \end{center}
\end{figure*}

WCSimAnalysis is a Water Cherenkov reconstruction package developed by Andrew Blake for the Long Baseline Neutrino Experiment (LBNE collaboration) \cite{Andy}. It provides a framework for generic event cleaning, track reconstruction, and particle id, and it comes equipped with variety of pre-built algorithms. A collaboration between Iowa State, the University of Chicago, and Argonne National Laboratory has continued adding to the code, developing new track-fitting techniques for Water Cherenkov detectors based on advanced photosensors with sub-cm imaging capabilities and timing resolutions below 100 picoseconds \cite{fast_group_and_LAPPD}. In this paper, we use only the simple, low-level reconstruction tools native to the package, though in future works we hope to implement more sophisticated algorithms.

The results presented in this paper rely on a simple vertex reconstruction algorithm, commonly known as a "point fit" \cite{reference_Super_K}. It assumes that all of the scintillation and Cherenkov light is emitted from a single point in space-time (x0,y0,z0,t0). In actuality, the light is emitted along an extended, multi-scattered electron track. However, at the energies discussed in this paper, the extent of this track is small (a few cm) compared to the scale of the detector (R=6 meters) and thus typical photon transit distances.

The first step of the reconstruction process relies on exact numerical calculations of vertex candidates from quadruplets of hits. Given a single point source, we need four constraints to solve for the four unknowns of the vertex \cite{Michael Smy} This approach would provide an exact solution in the case of four prompt, unscattered photons originating from a common point. However, many of these randomly chosen quadruplets will produce anomalous solutions due to real world effects such as optical scattering, delayed emission, and deviations from the point-like geometry. Nonetheless, we found that any chosen subset of 80 quadruplets was a sufficiently large ensemble to assure that some solutions will be close to the true vertex.

Once a small set of vertex candidates have been solved for, we test the goodness of each candidate and select the one that best fits the full ensemble of photon hits. The goodness of fit is determined based on the distribution of an observable known as the "point time residual" [super K]. The point time residual is calculated by taking the difference between the measured time of a photon hit, and the predicted time of the hit, given its distance from the vertex hypothesis, the speed of light in the scintillator, and the hypothesized t0 of the event. The width of the time residual distribution over all hits is minimized when the hypothesized vertex is near the true vertex. Based on this figure of merit, we select the vertex with the narrowest time residual distribution from among the 80 candidates. This best candidate is used as the seed value for a MINUIT-based optimization [MINUIT], where the vertex is allowed to vary over a continuum of locations until the global optimum is found.

The direction of the electron track is determined by taking the centroid of all vectors pointing from the fitted vertex to the hits on the detector. Since the Cherenkov light his highly directional, and since our timing cut enhances the purity of the Cherenkov light in the sample, this calculation provides a good measure of the track direction. Future work will focus on more sophisticated algorithms, taking better advantage of the highly constrained Cherenkov geometry. For the purpose of this paper, the current method provides a very good proof-of-principle.

For the purpose of testing the reconstruction algorithm described above we use 1000 simulated electrons with the energy of 5~MeV. The electrons are similated at the center of the detector, $\vec{r}$ = (0,0,0), along the x-axis, $\frac{\vec{p}}{|\vec{p}|}$ = (1,0,0).

Figure~\ref{fig:reco} show vertex reconstruction. The vertex is reasonably well reconstructed around center of the detector, $\vec{r}$ = (0,0,0). The RMS of the distributions for all three reconstructed coordinates are smaller than 3.5~cm. The shift along x-axis is attributed to the fact that electron is moving towards that direction and has a finite track length while the reconstruction algorithm treats all photons as they were emitted from the same point.

The reconstruction of the direction is shown in Fig.~\ref{fig:reco}. For majority of the events we are able to reconsruct the direction along x-axis.

%
%This is Matt's text from July 17.
%
%WCSimAnalysis is a Water Cherenkov reconstruction package developed by Andrew Blake for the Long Baseline Neutrino Experiment (LBNE collaboration). It provides a framework for generic event cleaning, track reconstruction, and particle id, and it comes equipped with variety of pre-built algorithms. A collaboration between Iowa State, the University of Chicago, and Argonne National Laboratory has continued adding to the code, developing new track-fitting techniques for Water Cherenkov detectors built around the unique capabilities of microchannel plate photodetectors (MCPs): timing resolutions below 100 picoseconds and imaging resolutions below a centimeter. In this paper, we use only the simple, low-level reconstruction tools native to the package, though in future works we hope to implement more sophisticated techniques.

%The simplest approach to vertex reconstruction, used in this paper, is commonly referred to as a Òpoint fitÓ. It assumes that, to first order, the light was emitted a single point in space-time (x0,y0,z0,t0). In reality, the light is emitted along the extended, multi-scattered track of the electron. However, at the energies discussed in this paper, the extent of the electron track is small (a few cm) compared to the scale of the detector, which has a radius of 6 meters. It is also small compared with the vertex resolutions typical of experiments with this scale. As starting point, the assumption works well.

%The first algorithm we use to find a seed vertex relies on exact numerical calculations of possible vertex candidates. Given a single point source, we need four constraints to solve for the four unknowns of the vertex. A collection of N random quadruplets of hits is selected and used to solve for N corresponding vertex candidates. Some of these solutions will produce nonsensical results due to optical scattering of the emitted light, slow decay constant of the scintillation, finite detector resolution, and the limitations of our assumption of a point source. Nonetheless, over the N unique quadruplets, we expect some of the solutions to lie near the true event vertex. In this paper we chose N to be 80, determined by trial and error. However, given many hundreds of hits, the number of possible 4-hit combinations is much larger and one can choose a different N to better suit oneÕs particular experimental context.

%Once the 80 vertex candidates have been solved for, we test the goodness of each candidate and select the one that best fits the full ensemble of hits. Distances are calculated between each hit and the candidate point-source. From these distances, we calculate the transit time, given the speed of light in our scintillator medium. This transit time plus the t0 of the 4-vertex provides a predicted arrival time for each hit. The Òtime residualÓ is the calculated as difference between the predicted and the measured arrival time, given the vertex candidate. Since we assume a point-source in this paper, we use what we call the Òsimple time residualÓ, as distinct from the Òextended time residualÓ, which assumes an extended, track-like geometry. The distribution of time residuals for each hit with respect to a vertex candidate has a width that is minimized when the hypothesis vertex is close to the true vertex. The t0 of the vertex candidate is left to float, and is adjusted to maximize the likelihood comparison with a Gaussian test function. The width of that gaussian test function is hand optimized to correspond with the width of the Òbest fitÓ. The vertex candidate with the largest maximum is selected to serve as the best candidate seed vertex. Once a single best seed vertex has been chosen, the same time-residual figure-of-merit is used in a MINUIT-based fit over a continuum of possible vertices in the neighborhood of the seed.

%Once the vertex fitting is completed, we determine the direction by taking the centroid of all vectors pointing from the fitted event vertex and the hits on the detector. Since our initial time cut greatly enhances the purity of the Cherenkov light in this sample of hits and, since the Cherenkov light is directional (compared to the isotropic scintillation light), this fit provides a good measure of the track direction. This approach does not rely on the highly constrained Cherenkov geometry of the event, which contains further information. Future work will focus on better algorithms. Nonetheless, for a proof-of-principles, this method works sufficiently well..
